### 2024-10-04 Paper/MSEPM-Manuscript-Updated.tex
Update MSEPM-Manuscript-Updated.tex

\abstract{Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging associated biology. 
Despite their widespread use, these methods do not account for other factors that also contribute to the variability of DNA methylation data.
 For example, many CpG sites show strong sex-specific or cell type specific patterns that likely impact the predictions of epigenetic age.  
 To overcome these limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic Pacemaker (MSEPM). 
 We show that the MSEPM is capable of accurately modeling multiple methylation associated factors simultaneously, while also providing site specific 
 models that describe the per site relationship between methylation and these factors. We utilized the MSEPM with a large aggregate cohort of blood
  methylation data to construct models of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models 
  capture a large faction of the variability at thousands of DNA methylation sites.  Moreover, this approach allows us to identify sites that 
  are primarily affected by aging and no other factors.  An analysis of these sites reveals that those that lose methylation over time are 
  enriched for CTCF transcription factor chip peaks, while those that gain methylation over time are associated with bivalent promoters of 
  genes that are not expressed in blood. These observations suggest mechanisms that underlie age associated methylation changes and suggest 
  that age associated increases in methylation may not have strong functional consequences on cell states. In conclusion, the MSEPM is 
  capable of accurately modeling multiple methylation associated factors and the models produced can illuminate site specific 
  combinations of factors that affect methylation dynamics.}

@@

We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. We simulated 1000 samples with the 
four epigenetic factors described above. We then simulated methylation values using the simulated site rates. Simulated samples
 were then split for training ($n=500$) and testing ($n=500$). MSEPM models were then fitted using the values of the input factors, $p^0_{k,j}$. 
 We generated 1000 simulated datasets and fit MSEPM models using four combinations of input factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). 
 Within each simulation, epigenetic state predictions and methylation site predictions were made for all testing samples. All models captured 
 the nonlinear association between simulated age and methylation (Supp. Figure 1). As the number of factors in the model is increased the mean 
 absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases (Figure 2A). Importantly, to accurately
  assess simulated age it is necessary to account for the influence of the other simulated factors (Sex, CN, CE).

@@

\begin{center}
    \begin{figure}
    \includegraphics[scale=.25]{Figures/Figure1.png}    
    \footnotesize
    \caption*{\small \textbf{Figure 1:}Simulated factors and the association with simulated methylation values. (A) Age with a non-linear association with methylation 
(B). Sex (C) with a binary association with methylation (D). Normal factor (E) with a linear relationship with methylation (F). Continuous exponential
 trait (G) with a linear relationship with methylation. (I) Simulated methylation sites. Each simulation site has a starting methylation value $r^0_i$ ,
  rate of change associated with each simulated factor $r_{i, factor}$ and range of variation $\delta_i$.}
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.2]{Figures/Figure2.png}
    \footnotesize
    \caption*{\small \textbf{Figure 2:} (A) The MAE of the factor predictions on the testing set as multiple factors are modeled simultaneously and 
    (B) predicted methylation MAE as factors are included in the MSEPM model where the centerline is the 50th quantile and the box 
    with greatest width contains 50\% of the underlying data with each smaller box containing 50\% of the remaining data 
    with 6 levels of box width. (C) Model coefficients for 
Age, Sex, Continuous Normal and Continuous Exponential factors for models trained $(n = 500)$ with all four simulated 
factors. (D). Simulated and predicted methylation values for all simulated testing sites across all training fold }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.2]{Figures/Figure3.png}
    \footnotesize
    \caption*{\small \textbf{Figure 3:} Distribution of age (A) and (B) sex in aggregate blood dataset. (C) Calculated cell type composition and (D) loading plot of
 principal components of cell type composition in aggregate blood data set.}
    \end{figure}
\end{center}

\begin{raggedleft}
\begin{figure}
\includegraphics[scale=.15]{Figures/Figure4.png}
\footnotesize
\caption*{\small \textbf{Figure 4:} MSEPM model trained with age, sex, CT-PC1 and CT-PC2 predictions within testing 
set for epigenetic factors (A) age, (B) sex, (C) CT-PC1 and (D) CT-PC2. (E) Observed and predicted
 methylation values for training set has high concordance $(R^2=0.933, MAE=0.035)$}
\end{figure}
\end{raggedleft}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.15]{Figures/Figure5.png}
    \footnotesize
    \caption*{\small \textbf{Figure5: } (A) Site clustering by standardized model coefficients. Sites clusters show 
    distinct relationships with modeled traits. (B-K) Top five enriched transcription factors for clusters 1 - 10. 
    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.15]{Figures/Figure6.png}
    \footnotesize
    \caption*{\small \textbf{Figure6: } (A) Heatmap of H3K4me3 ChIP-seq enrichment for two different blood cell types (CD38\textsuperscript{+} B Cells and CD56\textsuperscript{+} NTK Cells) in 
two cohorts of individual within 0 to 5 years old and 60 to 65 years old. The average level within 2kb up and downstream for centered genomic regions of cluster 7 is 
represented above the heatmap. (B) Genome browser view of H3K4me3 levels in each cohort at the promoter regions of \textit{KCTD1} and \textit{IRS2} genes.  
    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.15]{Figures/Figure7.png}
    \footnotesize
    \caption*{\small \textbf{Figure7: } (A) Heatmap of H3K4me3 ChIP-seq enrichment for two different blood cell types (CD38\textsuperscript{+} B Cells and CD56\textsuperscript{+} NTK Cells) in 
two cohorts of individual within 0 to 5 years old and 60 to 65 years old. The average level within 2kb up and downstream for centered genomic regions of cluster 7 is 
represented above the heatmap. (B) Genome browser view of H3K4me3 levels in each cohort at the promoter regions of \textit{KCTD1} and \textit{IRS2} genes. }
    \end{figure}
\end{center}

>>>>

\section{Abstract}\label{abstract}

Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging associated biology. 
Despite their widespread use, these methods do not account for other factors that also contribute to the variability of DNA methylation data.
 For example, many CpG sites show strong sex-specific or cell type specific patterns that likely impact the predictions of epigenetic age.  
 To overcome these limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic Pacemaker (MSEPM). 
 We show that the MSEPM is capable of accurately modeling multiple methylation associated factors simultaneously, while also providing site specific 
 models that describe the per site relationship between methylation and these factors. We utilized the MSEPM with a large aggregate cohort of blood
  methylation data to construct models of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models 
  capture a large faction of the variability at thousands of DNA methylation sites.  Moreover, this approach allows us to identify sites that 
  are primarily affected by aging and no other factors.  An analysis of these sites reveals that those that lose methylation over time are 
  enriched for CTCF transcription factor chip peaks, while those that gain methylation over time are associated with bivalent promoters of 
  genes that are not expressed in blood. These observations suggest mechanisms that underlie age associated methylation changes and suggest 
  that age associated increases in methylation may not have strong functional consequences on cell states. In conclusion, the MSEPM is 
  capable of accurately modeling multiple methylation associated factors and the models produced can illuminate site specific 
  combinations of factors that affect methylation dynamics.

@@

We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. We simulated 1000 samples with the 
four epigenetic factors described above. We then simulated methylation values using the simulated site rates. Simulated samples
 were then split for training ($n=500$) and testing ($n=500$). MSEPM models were then fitted using the values of the input factors, $p^0_{k,j}$. 
 We generated 1000 simulated datasets and fit MSEPM models using four combinations of input factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). 
 Within each simulation, epigenetic state predictions and methylation site predictions were made for all testing samples. All models captured 
 the nonlinear association between simulated age and methylation (Supp. Figure 1). As the number of factors in the model is increased, the mean 
 absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases (Figure 2A). Importantly, to accurately
  assess simulated age it is necessary to account for the influence of the other simulated factors (Sex, CN, CE).

@@

\begin{center}
    \begin{figure}
    \includegraphics[scale=.25]{Figure1.png}    
    \footnotesize
    \caption*{\small \textbf{Figure 1:}Simulated factors and the association with simulated methylation values. (A) Age with a non-linear association with methylation 
(B). Sex (C) with a binary association with methylation (D). Normal factor (E) with a linear relationship with methylation (F). Continuous exponential
 trait (G) with a linear relationship with methylation. (I) Simulated methylation sites. Each simulation site has a starting methylation value $r^0_i$ ,
  rate of change associated with each simulated factor $r_{i, factor}$ and range of variation $\delta_i$.}
    \end{}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.2]{Figure2.png}
    \footnotesize
    \caption*{\small \textbf{Figure 2:} (A) The MAE of the factor predictions on the testing set as multiple factors are modeled simultaneously and 
    (B) predicted methylation MAE as factors are included in the MSEPM model where the centerline is the 50th quantile and the box 
    with greatest width contains 50\% of the underlying data with each smaller box containing 50\% of the remaining data 
    with 6 levels of box width. (C) Model coefficients for 
Age, Sex, Continuous Normal and Continuous Exponential factors for models trained $(n = 500)$ with all four simulated 
factors. (D). Simulated and predicted methylation values for all simulated testing sites across all training fold }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.2]{Figure3.png}
    \footnotesize
    \caption*{\small \textbf{Figure 3:} Distribution of age (A) and (B) sex in aggregate blood dataset. (C) Calculated cell type composition and (D) loading plot of
 principal components of cell type composition in aggregate blood data set.}
    \end{figure}
\end{center}

\begin{raggedleft}
\begin{figure}
\includegraphics[scale=.15]{Figure4.png}
\footnotesize
\caption*{\small \textbf{Figure 4:} MSEPM model trained with age, sex, CT-PC1 and CT-PC2 predictions within testing 
set for epigenetic factors (A) age, (B) sex, (C) CT-PC1 and (D) CT-PC2. (E) Observed and predicted
 methylation values for training set has high concordance $(R^2=0.933, MAE=0.035)$}
\end{figure}
\end{raggedleft}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.15]{Figure5.png}
    \footnotesize
    \caption*{\small \textbf{Figure5: } (A) Site clustering by standardized model coefficients. Sites clusters show 
    distinct relationships with modeled traits. (B-K) Top five enriched transcription factors for clusters 1 - 10. 
    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.15]{Figure6.png}
    \footnotesize
    \caption*{\small \textbf{Figure6: } (A) Heatmap of H3K4me3 ChIP-seq enrichment for two different blood cell types (CD38\textsuperscript{+} B Cells and CD56\textsuperscript{+} NTK Cells) in 
two cohorts of individual within 0 to 5 years old and 60 to 65 years old. The average level within 2kb up and downstream for centered genomic regions of cluster 7 is 
represented above the heatmap. (B) Genome browser view of H3K4me3 levels in each cohort at the promoter regions of \textit{KCTD1} and \textit{IRS2} genes.  
    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.15]{Figure7.png}
    \footnotesize
    \caption*{\small \textbf{Figure7: } (A) Heatmap of H3K4me3 ChIP-seq enrichment for two different blood cell types (CD38\textsuperscript{+} B Cells and CD56\textsuperscript{+} NTK Cells) in 
two cohorts of individual within 0 to 5 years old and 60 to 65 years old. The average level within 2kb up and downstream for centered genomic regions of cluster 7 is 
represented above the heatmap. (B) Genome browser view of H3K4me3 levels in each cohort at the promoter regions of \textit{KCTD1} and \textit{IRS2} genes. }
    \end{figure}
\end{center}
